---
title: "EDA_Yuqi"
author: "Yuqi Zhang"
date: "2024-05-14"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(data.table)
library(ggplot2)
library(dplyr)
library(GGally)
library(devtools)
#install_github("mfasiolo/electBook")
library(electBook)
library(rjags)
library(R2jags)
#install.packages("INLA",repos=c(getOption("repos"),INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)
library(INLA)
library(bsts)
```



```{r}
load("Irish.RData")
str(Irish)
summary(Irish)
```


```{r}
demand_data <- as.data.frame(Irish$indCons)

mean_demand <- colMeans(demand_data)
mean_demand_df <- data.frame(ID = colnames(demand_data), mean_demand = mean_demand)


survey_data <- Irish$survey
survey_data <- merge(survey_data, mean_demand_df, by.x = "ID", by.y = "ID")
```

The mean demand is a straightforward measure that represents the average electricity consumption of each household over the observation period. By providing a single value per household, it facilitates easy comparison and understanding of overall consumption patterns. Mean demand captures general consumption behaviour, smoothing out short-term fluctuations and anomalies. This measure is effective for identifying major patterns and differences between households.

```{r}
survey_data$HEAT.WATER <- as.factor(survey_data$HEAT.WATER)
survey_data$WINDOWS.doubleglazed <- as.factor(survey_data$WINDOWS.doubleglazed)
survey_data$ResTariffallocation <- as.factor(survey_data$ResTariffallocation)
survey_data$ResStimulusallocation <- as.factor(survey_data$ResStimulusallocation)

# Boxplots of mean demand by categorical variables
ggplot(survey_data, aes(x = HEAT.WATER, y = mean_demand)) +
  geom_boxplot() +
  labs(title = "Mean Demand by HEAT.WATER", x = "HEAT.WATER", y = "Mean Demand")

# Double glazed window, 5 levels
ggplot(survey_data, aes(x = WINDOWS.doubleglazed, y = mean_demand)) +
  geom_boxplot() +
  labs(title = "Mean Demand by WINDOWS.doubleglazed", x = "WINDOWS.doubleglazed", y = "Mean Demand")

ggplot(survey_data, aes(x = ResTariffallocation, y = mean_demand)) +
  geom_boxplot() +
  labs(title = "Mean Demand by ResTariffallocation", x = "ResTariffallocation", y = "Mean Demand")

#residential tariffs allocation
ggplot(survey_data, aes(x = ResStimulusallocation, y = mean_demand)) +
  geom_boxplot() +
  labs(title = "Mean Demand by ResStimulusallocation", x = "ResStimulusallocation", y = "Mean Demand")

# Scatter plots for numerical variables
ggplot(survey_data, aes(x = HOME.APPLIANCE..White.goods., y = mean_demand)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Mean Demand by HOME.APPLIANCE..White.goods.", x = "Number of Home Appliances", y = "Mean Demand")

ggplot(survey_data, aes(x = Code, y = mean_demand)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Mean Demand by Code", x = "Code", y = "Mean Demand")

```


```{r}
# Visualize the daily demand patterns for a sample of households
library(ggplot2)
set.seed(123)
sample_households <- sample(ncol(demand_data), 5)

# Convert to long format for ggplot
demand_sample <- demand_data[, sample_households]
demand_long <- data.frame(time = rep(1:nrow(demand_sample), ncol(demand_sample)),
                          demand = as.vector(as.matrix(demand_sample)),
                          household = rep(names(demand_sample), each = nrow(demand_sample)))

ggplot(demand_long, aes(x = time, y = demand, color = household)) +
  geom_line() +
  labs(title = "Electricity Demand for Sample Households", x = "Time", y = "Demand")

```

```{r}
# Group households based on the number of home appliances
survey_data_grouped <- survey_data %>%
  group_by(HOME.APPLIANCE..White.goods.) %>%
  summarize(mean_demand = mean(mean_demand, na.rm = TRUE),
            .groups = 'drop')
```


# Define a Bayesian Model

```{r}
# Prepare data for JAGS
jags_data <- list(
  N = nrow(survey_data_grouped),
  appliances = survey_data_grouped$HOME.APPLIANCE..White.goods.,
  demand = survey_data_grouped$mean_demand
)

# Define the JAGS model
jags_model <- "
model {
  for (i in 1:N) {
    demand[i] ~ dnorm(mu[i], tau)
    mu[i] <- beta0 + beta1 * appliances[i]
  }
  
  beta0 ~ dnorm(0, 0.001)
  beta1 ~ dnorm(0, 0.001)
  tau ~ dgamma(0.001, 0.001)
  sigma <- 1 / sqrt(tau)
}
"

```


```{r}
# Initial values for MCMC
init_values <- function() {
  list(beta0 = rnorm(1), beta1 = rnorm(1), tau = rgamma(1, 1, 1))
}

# Parameters to monitor
params <- c("beta0", "beta1", "sigma")

# Run the JAGS model
jags_fit <- jags(data = jags_data, 
                 inits = init_values, 
                 parameters.to.save = params, 
                 model.file = textConnection(jags_model), 
                 n.chains = 3, 
                 n.iter = 2000, 
                 n.burnin = 1000, 
                 n.thin = 2)

# Print the results
print(jags_fit)

```






