---
title: "Report"
author: "Xin Guan, Vera Hudak, Yuqi Zhang"
date: "2024-05-24"
output: html_document
---
```{r}
library(electBook)
library(dplyr)
library(tidyr)
library(lubridate)
library(proxy)
library(tibble)
library(ggplot2)
```

# Modeling Household Electricity Demand

## Introduction



## Explanatory Data Analysis

Loading Data:
```{r}
load("Irish.RData")
```

The file `Irish` contains three data sets, the data set`Irish$indCons` is the 
electricity demand for 2671 household, with a resolution of half hour over the 
time period 2019-12-29 to 2020-12-29.
```{r}
head(Irish$indCons[,1:10])
```
The data set `Irish$extra` contains variables such as `tod` (time of day), 
`toy` (time of year), `dow` (day of the week), and more.
```{r}
head(Irish$extra)
```

```{r}
head(Irish$survey)
```


```{r}
# Count zeros in each row
row_zero_counts <- rowSums(Irish$indCons == 0)

hist(row_zero_counts)
```


```{r}
sum(apply(Irish$indCons, 2, function(x) sum(x == 0)))
#There are many zeros in the demand data frame
print(16799*2674)
```

```{r}
# Count zeros in each column
col_zero_counts <- colSums(Irish$indCons == 0)

# Histogram of columns with more than 100 zeros
hist(col_zero_counts[col_zero_counts > 100], 
     main="Histogram of Households with More Than 100 Zero Usage", 
     xlab= "Zero Counts per Household")
```

```{r}
sum(col_zero_counts > 30*48)
```

```{r}
cols_to_remove <- which(col_zero_counts > 30*48)
```

```{r}
df0 <- Irish$indCons[,-cols_to_remove]

df0$time_mean_dem <- rowSums(Irish$indCons)/ncol(Irish$indCons)
```



```{r}
df <- cbind(df0[,"time_mean_dem"],Irish$extra)
colnames(df) <- c("time_mean_demand", colnames(Irish$extra))
```



```{r}
head(df)
```

## Visualizing main characteristics

```{r}
# Basic summary of each column
summary(df)
#`holy` is all FALSE

# Time series plot of time_mean_demand
ggplot(df, aes(x=dateTime, y=time_mean_demand)) + geom_line() + 
  ggtitle("Time Series of Mean Demand")

# Boxplots to check variation of time_mean_demand across days of the week
ggplot(df, aes(x=dow, y=time_mean_demand)) + geom_boxplot() + 
  ggtitle("Demand Variation by Day of Week")

```

```{r}
# Scatter plot of time_mean_demand vs. temperature
ggplot(df, aes(x=temp, y=time_mean_demand)) +
    geom_point(alpha=0.5) +
    geom_smooth(method="lm", se=FALSE, color="blue") +
    labs(x="Temperature", y="Mean Demand", 
         title="Relationship Between Temperature and Mean Demand")
```


```{r}
# Line plot for time_mean_demand across different times of day
ggplot(df, aes(x=tod, y=time_mean_demand, group=1)) +
    geom_point(alpha=0.5) +
    geom_smooth(color="blue") +
    labs(x="Time of Day", y="Sum Demand", 
         title="Mean Demand Across Different Times of Day")

```


## Aggregation


```{r}
# Load the dataset
data(Irish)

# Calculate the number of zero values in each column
col_zero_counts <- colSums(Irish$indCons == 0)

# Identify columns to remove (columns with more than 30*48 zero values)
cols_to_remove <- which(col_zero_counts > 30 * 48)

# Create a data frame with demand data and remove identified columns
df <- Irish$indCons
df <- df[,-cols_to_remove]

# Add date and time columns
df$date <- as.Date(Irish$extra$dateTime)
df$time <- format(Irish$extra$dateTime, "%H:%M:%S")

# Gather the data into long format
df_long <- df %>%
  pivot_longer(cols = -c(date, time), names_to = "household_id", 
               values_to = "demand")

# Calculate the average demand over the year for each 30-minute interval
avg_demand <- df_long %>%
  group_by(household_id, time) %>%
  summarise(average_demand = mean(demand, na.rm = TRUE)) %>%
  ungroup()

avg_demand_wide <- avg_demand %>%
  pivot_wider(names_from = time, values_from = average_demand)

# Display the result
print(avg_demand_wide)
```


```{r}
# Compute the cosine similarity matrix
compute_cosine_similarity_matrix <- function(data) {
  data_matrix <- as.matrix(data[-1])  # Remove the household_id column
  similarity_matrix <- proxy::simil(data_matrix, method = "cosine")
  dist_matrix <- 1 - similarity_matrix
  return(as.matrix(dist_matrix))
}

cosine_distances <- compute_cosine_similarity_matrix(avg_demand_wide)

# Hierarchical clustering
hc <- hclust(as.dist(cosine_distances), method = "ward.D2")

# Plot the dendrogram
plot(hc, labels = FALSE, main = "Dendrogram of Households", xlab = "Households", 
     ylab = "Height")
abline(h=1,col="red")
```

```{r}
# Create clusters
clusters <- cutree(hc, k = 6)
avg_demand_wide$cluster <- clusters
# Summarize the number of households in each cluster
cluster_summary <- avg_demand_wide %>%
  group_by(cluster) %>%
  summarise(num_households = n())

# Display the summary
print(cluster_summary)

# Reshape avg_demand back to long format
avg_demand_long <- avg_demand_wide %>%
  pivot_longer(cols = -c(household_id, cluster), names_to = "time", 
               values_to = "daily_demand")


# Join cluster information back to the original dataframe
df_with_clusters <- df_long %>%
  left_join(avg_demand_long, by = c("household_id", "time"))

# Analyze cluster characteristics
cluster_analysis <- df_with_clusters %>%
  group_by(cluster) %>%
  summarise(
    average_demand = mean(daily_demand, na.rm = TRUE)
  )

print(cluster_analysis)
```


```{r}
df_t <- as.data.frame(t(df[,-c(ncol(df)-1,ncol(df))]))

# Step 2: Add the clusters as a new column to the transposed data frame
df_t$cluster <- clusters

# Step 3: Group by cluster and calculate the mean for each row within each 
#cluster
mean_by_cluster <- df_t %>%
  group_by(cluster) %>%
  summarise(across(everything(), mean, na.rm = TRUE))

mean_by_cluster <- as.data.frame(mean_by_cluster[,-1])
rownames(mean_by_cluster) <- c("Cluster 1","Cluster 2","Cluster 3","Cluster 4",
                               "Cluster 5","Cluster 6")
# View the result
print(mean_by_cluster)
```

```{r}
df0 <- Irish$extra
df0 <- df0 %>% mutate(dow = ifelse(dow %in% c("Sat", "Sun"), "True", "False")) 
%>% select(-c(holy,time,dateTime))%>%rename(weekend = dow)
df0 <-t(df0)
```

```{r}
colnames(df0) <- colnames(mean_by_cluster)
mean_by_cluster <- rbind(mean_by_cluster,df0)
mean_by_cluster <- as.data.frame(t(mean_by_cluster))
# Convert Cluster  and temp columns to numeric, weekend into logic
mean_by_cluster <- mean_by_cluster %>%
  mutate(across(starts_with("Cluster"), as.numeric),
         weekend = as.logical(weekend),
         temp = as.numeric(temp))
```

```{r}
# Save as CSV file
#write.csv(mean_by_cluster, file = "AggregatedData1.csv", row.names = FALSE)
```



```{r}
avg_demand_wide$cluster <- clusters
```


```{r}
#write.csv(avg_demand_wide, file = "cluster_data.csv", row.names = FALSE)
```



```{r}
# Identify the time columns by excluding non-time columns
time_columns <- grep("^\\d{2}:\\d{2}:\\d{2}$", colnames(avg_demand_wide), 
                     value = TRUE)

# Add rowid to avg_demand_wide for sampling
avg_demand_wide <- avg_demand_wide %>%
  mutate(rowid = row_number())

# Convert data to long format for plotting
avg_demand_long <- avg_demand_wide %>%
  pivot_longer(cols = all_of(time_columns), 
               names_to = "Time", 
               values_to = "Demand") %>%
  mutate(Time = as.numeric(gsub(":", "", Time)))

# Calculate mean demand for each cluster
mean_demand <- avg_demand_long %>%
  group_by(cluster, Time) %>%
  summarize(mean_demand = mean(Demand), .groups = 'drop')

# Plot mean demand for each cluster
for (cl in unique(avg_demand_wide$cluster)) {
  # Filter data for the cluster
  cluster_data <- avg_demand_long %>% filter(cluster == cl)
  
  # Sample 10 houses from the cluster
  sampled_houses <- sample(unique(cluster_data$rowid), 10)
  sample_data <- cluster_data %>% filter(rowid %in% sampled_houses)
  
  # Plot mean demand
  p <- ggplot() +
    geom_line(data = mean_demand %>% filter(cluster == cl), 
              aes(x = Time, y = mean_demand, color = "Mean Demand"), size = 1) +
    geom_line(data = sample_data, 
              aes(x = Time, y = Demand, group = rowid, 
                  color = as.factor(rowid)), alpha = 0.3) + 
    labs(title = paste("Cluster", cl),
         x = "Time",
         y = "Demand") +
    scale_color_manual(values = c("Mean Demand" = "red", setNames(rainbow(10), 
                                                            sampled_houses))) +
    theme_minimal()
  
  print(p)
}

```























